---
title: "HW1 Markdown"
author: "Caden Zonnefeld"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: pdf_document
---

Changes made ot this file at 12:15 PM

```{r, include=FALSE}
knitr::opts_chunk$set(results = "hide")
```
 
## Question 1
I assigned the iris data to a variable, added an ID column, and created a simple sumRow function that calculates the sum of numeric entries.
```{r}
myData <- iris
myData$id <- 1:nrow(myData)
sumRow <- function(x,i){
  numericColumns <- x[1:4]
  sum(numericColumns[i,])
}
```

## Question 2
I used a for loop to calculate the row sums of the numeric data entries.
```{r}
numericColumns <- myData[1:4]
rowSums <- vector(length=150L)
for (i in 1:150){
  rowSums[i] <- sum(numericColumns[i,])
}
```

## Question 3
I wrote another version of a sum row function, this time to be used by the apply function.
```{r}
rowSums2 <- vector(length=150L)
sumRow2 <- function(x){
  numericColumns <- x[1:4]
  sum(numericColumns)
}
rowSums2 <- apply(myData[1:4],1,sumRow2)
```

## Question 4
I modified my sum row function so that it could be used with the sapply function, this required me to first create a transposed version of the dataset.
```{r}
rowSums3 <- vector(length=150L)
transposeData <- data.frame(t(myData[1:4]))
typeof(transposeData)
sumRow3 <- function(x){
  sum(x)
}
rowSums3 <- sapply(transposeData,sumRow3)
```

## Question 5
Using the tidyr library, I converted the data into a long form of dataset.
```{r}
library(tidyr)
longData <- gather(myData, key = "FlowerAttributes", value = "Values", Sepal.Length,
                   Sepal.Width, Petal.Length, Petal.Width)
```

## Question 6
I used the aggregate function to find the mean value for each flower attribute.
```{r}
aggregateNumeric <- aggregate(Values~FlowerAttributes,longData,mean)
aggregateNumeric["AverageValueForAttribute"] <- aggregateNumeric["Values"]
aggregateNumeric <- aggregateNumeric[-2]
```

## Question 7
I merged the aggregated mean values from the previous question with the long dataset.
```{r}
longData <- merge(longData,aggregateNumeric,by="FlowerAttributes")
```

## Question 8
I repeated the process of mean aggregation and merging, this time using the mean for each flower species.
```{r}
aggregateSpecies <- aggregate(Values~Species+FlowerAttributes,longData,mean)
aggregateSpecies["AverageValueForSpecies"] <- aggregateSpecies["Values"]
aggregateSpecies <- aggregateSpecies[-3]

completeLongData <- merge(longData,aggregateSpecies,by=c("Species","FlowerAttributes"))
```

## Question 9
I reversed the data from a long format into a wide format.
```{r, message=FALSE}
library(reshape2)
wideData <- dcast(completeLongData,formula=Species+AverageValueForSpecies+
                    AverageValueForAttribute+id~FlowerAttributes,value.var="Values")
wideData <- wideData[-4]
```

## Question 10
Set the working directory, loaded in each of the files, and set widths for each file.
```{r}
setwd("Documents/STAT 344")
ia97 <- data.frame(read.fwf("1997ia.txt",widths=c(3,4,9,9,9,5,5,5,9,9,9,5,5,5,9,9,9,
                                                  5,5,5,7,7,7,8,8,8,5,5,5,46,3)))
ia18 <- data.frame(read.fwf("2018ia.txt",widths=c(3,4,9,9,9,5,5,5,9,9,9,5,5,5,9,9,9,
                                                  5,5,5,7,7,7,8,8,8,5,5,5,46,3,22)))
```
Added a year column to each of the datasets.
```{r}
ia97["Year"] <- 1997
ia18["Year"] <- 2018
```
Renamed the columns in each of the datasets so that they describe their contents.
```{r, message=FALSE}
library(dplyr)

ia97 <- ia97 %>%
  rename(StateCode = V1, CountyCode = V2, PovEstimate = V3, PovEstimate90CILower = V4,
         PovEstimate90CIUpper = V5, PovertyPerc = V6, PovertyPerc90CILower = V7,
         PovertyPerc90CIUpper = V8, PovEstimate0to17 = V9,
         PovEstimate0to1790CILower = V10,
         PovEstimate0to1790CIUpper = V11, PovEstimatePerc0to17 = V12,
         PovEstimatePerc0to1790CILower = V13, PovEstimatePerc0to1790CIUpper = V14, 
         EstimateChild5to17Pov = V15, EstimateChild5to17Pov90CILower = V16,
         EstimateChild5to17Pov90CIUpper = V17, EstimatedChild5to17PovPerc = V18,
         EstimatedChild5to17PovPerc90CILower = V19,
         EstimatedChild5to17PovPerc90CIUpper = V20, MedianHouseholdIncome = V21,
         MedianHouseholdIncome90CILower = V22, MedianHouseholdIncome90CIUpper = V23,
         EstimateChildUnder5Pov = V24, EstimateChildUnder5Pov90CILower = V25,
         EstimateChildUnder5Pov90CIUpper = V26, EstimatedChildUnder5PovPerc = V27,
         EstimatedChildUnder5PovPerc90CILower = V28,
         EstimatedChildUnder5PovPerc90CIUpper = V29, StateOrCountyName = V30,
         PostalStateCode = V31)
ia18 <- ia18 %>%
  rename(StateCode = V1, CountyCode = V2, PovEstimate = V3, PovEstimate90CILower = V4,
         PovEstimate90CIUpper = V5, PovertyPerc = V6, PovertyPerc90CILower = V7,
         PovertyPerc90CIUpper = V8, PovEstimate0to17 = V9,
         PovEstimate0to1790CILower = V10,
         PovEstimate0to1790CIUpper = V11, PovEstimatePerc0to17 = V12,
         PovEstimatePerc0to1790CILower = V13, PovEstimatePerc0to1790CIUpper = V14, 
         EstimateChild5to17Pov = V15, EstimateChild5to17Pov90CILower = V16,
         EstimateChild5to17Pov90CIUpper = V17, EstimatedChild5to17PovPerc = V18,
         EstimatedChild5to17PovPerc90CILower = V19,
         EstimatedChild5to17PovPerc90CIUpper = V20, MedianHouseholdIncome = V21,
         MedianHouseholdIncome90CILower = V22, MedianHouseholdIncome90CIUpper = V23,
         EstimateChildUnder5Pov = V24, EstimateChildUnder5Pov90CILower = V25,
         EstimateChildUnder5Pov90CIUpper = V26, EstimatedChildUnder5PovPerc = V27,
         EstimatedChildUnder5PovPerc90CILower = V28,
         EstimatedChildUnder5PovPerc90CIUpper = V29, StateOrCountyName = V30,
         PostalStateCode = V31, FileName = V32)
```
Added a margin of error column to each dataset.
```{r}
ia97["MarginOfError"] <- abs(ia97$PovEstimate90CIUpper-ia97$PovEstimate)
ia18["MarginOfError"] <- abs(ia18$PovEstimate90CIUpper-ia18$PovEstimate)
```
Made a subset of the 1997 and 2018 data for the suggested variables of year, county, poverty estimate,
margin of error for the poverty estimate, and poverty percentage estimate.
```{r}
ia97trimmed <- subset(ia97, select=c("Year","StateOrCountyName","PovEstimate",
                                     "MarginOfError","PovertyPerc"))
ia18trimmed <- subset(ia18, select=c("Year","StateOrCountyName","PovEstimate",
                                     "MarginOfError","PovertyPerc"))
```
Utilized an rbind to combine the data.
```{r}
cleanData <- rbind(ia97trimmed,ia18trimmed)
```
Renamed the columns for clarity and added the year for each measure so that year columns could be dropped.
```{r}
cleanData <- cleanData %>%
  rename(County = StateOrCountyName, PovertyEstimate = PovEstimate,
         MarginOfError = MarginOfError, PovertyPercentageEstimate = PovertyPerc)
```
Wrote the data into a comma separated value file.
```{r}
write.csv(cleanData,file="cleanSAIPEdata.csv")
```





